---
title: "EV Power - Lab 4 Project Report"
format: typst
---

# Example Solution 1

## **Part 0: libraries**

```{r}
library(tidyverse) 
library(janitor) 
library(sf) 
library(viridis)
library(readxl)
library(kableExtra)
library(leaflet)
library(tigris)
library(gganimate)

```

## **Part 1:** **Defining Research Question**

Chosen Question:
The research question for this report was to determine the question that Electric vehicles reduce direct emissions, but does the electricity used to charge them actually come from
clean sources? Therefore, one of the sub-questions would be to determine how the share of renewable energy has changed from 2021 to 2023. The other question would be to determine the proportion of electricity used in each state that comes from clean sources between 2021 and 2023. The other sub-question would be to determine whether the states with a higher share of renewable electricity are also the ones with more EV registrations in 2023. Finally, the other sub-question would be to identify how much renewable electricity is available to registered EVs in each state.
## Data and Methods
The following datasets were used to carry out the needed analysis;

      1. renew_use_2021.csv, 
      2. renew_use_2022.csv, 
      3. renew_use_2023.csv	2021–2023	
      4. total_energy_use_2021.csv, 
      5. total_energy_use_2022.csv, 
      6. total_energy_use_2023.csv	
      7. ev-registrationsby_state_2023.csv

## **Part 2: Data Preparation and Cleaning**

```{r}
data_path <- "C:/Users/ThinkPad/Desktop/Project 4/"

# Load and clean renewable and total energy data
renew21 <- read_csv(paste0(data_path, "renew-use-2021.csv")) %>% clean_names() %>% mutate(year = 2021)
renew22 <- read_csv(paste0(data_path, "renew-use-2022.csv")) %>% clean_names() %>% mutate(year = 2022)
renew23 <- read_csv(paste0(data_path, "renew-use-2023.csv")) %>% clean_names() %>% mutate(year = 2023)

total21 <- read_csv(paste0(data_path, "total-use-2021.csv")) %>% clean_names() %>% mutate(year = 2021)
total22 <- read_csv(paste0(data_path, "total-use-2022.csv")) %>% clean_names() %>% mutate(year = 2022)
total23 <- read_csv(paste0(data_path, "total-use-2023.csv")) %>% clean_names() %>% mutate(year = 2023)
ev_file <- "C:/Users/ThinkPad/Desktop/Project 4/ev-registrations-by-state-2023.csv"
ev23 <- read_csv(ev_file) %>%
  clean_names() 
ev23 <- ev23 %>%
  mutate(state = str_to_title(state))
head(ev23)
# average price from 2021 to 2023
price_file <- "C:/Users/ThinkPad/Desktop/Project 4/av-energy-price-2021-2023.xlsx"
av_price <- read_excel(price_file) %>%
  clean_names() 
av_price <- av_price %>%
  mutate(state = str_to_title(state))
head(av_price)
# we create an average price for the three years as;
av_price <- av_price %>%
  mutate(
    avg_price = rowMeans(select(., av_price_2021, av_price_2022, av_price_2023), na.rm = TRUE)
  ) %>%
  select(state, avg_price)
head(av_price)
## cleaning
ev23 <- ev23 %>%
  mutate(
    count_e_vs = str_replace_all(count_e_vs, "[^0-9]", ""), 
    count_e_vs = as.numeric(count_e_vs)                       
  )
head(ev23)
Project4 <- read_csv("C:/Users/ThinkPad/Desktop/Project 4/Project4.csv")%>%
  clean_names()
glimpse(Project4)
summary(select(Project4,renewable_use_2021, renewable_use_2022,renewable_use_2023,total_renewable_energy_2021,total_renewables_energy_2022, total_renewable_energy_2023, av_price_2021,av_price_2022,av_price_2023, e_vs))
## cleaning renewable energy data
```

## **Part 3: Joining / Pivoting Datasets for Analysis**

```{r}
project_long <- Project4 %>%
  pivot_longer(
    cols = matches("renewable_use_|total_.*_202\\d|av_price_"),
    names_to = c(".value", "year"),
    names_pattern = "(.*)_(202\\d)"
  )
## This code cleans and converts variables 
project_long <- project_long %>%
  mutate(
    year = as.integer(year),
    renewable_use = as.numeric(str_replace_all(renewable_use, "[^0-9\\.]", "")),
    total_renewable_energy = as.numeric(str_replace_all(total_renewable_energy, "[^0-9\\.]", "")),
    av_price = as.numeric(str_replace_all(av_price, "[$]", "")),
    e_vs = as.numeric(e_vs)  # EV column
  ) %>%
  select(state_symbol, state, year, renewable_use, total_renewable_energy, av_price, e_vs)
## This code creates the required variables for analysis
project_long <- project_long %>%
  mutate(
    pct_renewable = (renewable_use / total_renewable_energy) * 100,
    ev_per_total_energy = e_vs / total_renewable_energy
  )
head(project_long)
library(kableExtra)
summary_table <- project_long %>%
  group_by(year) %>%
  summarise(
    Mean_Renewable_Use = round(mean(renewable_use, na.rm = TRUE), 2),
    SD_Renewable_Use = round(sd(renewable_use, na.rm = TRUE), 2),
    Mean_Total_Renewable_Energy = round(mean(total_renewable_energy, na.rm = TRUE), 2),
    Mean_Pct_Renewable = round(mean(pct_renewable, na.rm = TRUE), 2),
    Mean_EV_Ratio = round(mean(ev_per_total_energy, na.rm = TRUE), 4),
    Mean_Price = round(mean(av_price, na.rm = TRUE), 2),
    States_Observed = n()
  )
kable(summary_table, caption = "Summary Statistics of Energy and EV Indicators by Year")
```

## **Part 4: Mapping Visualization**

```{r}
states_sf <- tigris::states(cb = TRUE) %>% 
  st_as_sf() %>%
  filter(!STUSPS %in% c("AL","AK","AZ","AR","CA","CO","CT","DE","FL","GA","HI","ID","IL","IN","IA",
    "KS","KY","LA","ME","MD","MA","MI","MN","MS","MO","MT","NE","NV","NH","NJ",
    "NM","NY","NC","ND","OH","OK","OR","PA","RI","SC","SD","TN","TX","UT","VT",
    "VA","WA","WV","WI","WY","DC"))
project_long <- project_long %>%
  filter(year %in% c(2021, 2022, 2023))
energy_2023 <- project_long %>%
  filter(year == 2023) %>%
  mutate(state_name = str_to_title(state))

# This code will help in merging the desired states by state name
states_merged <- states_sf %>%
  left_join(energy_2023, by = c("NAME" = "state_name"))
states_sf <- states(cb = TRUE)  
states_sf <- states_sf[!states_sf$NAME %in% c("Puerto Rico", "Guam", 
                                               "American Samoa", 
                                               "Commonwealth of the Northern Mariana Islands", 
                                               "United States Virgin Islands"), ]
states_merged <- left_join(states_sf, project_long, by = c("NAME" = "state"))
energy_2023 <- project_long %>%
  filter(year == 2023)
states_merged <- left_join(states_sf, energy_2023, by = c("NAME" = "state"))
ggplot(states_merged) +
  geom_sf(aes(fill = pct_renewable), color = "white") +
  scale_fill_viridis_c(option = "plasma", na.value = "grey90") +
  labs(
    title = "Percentage of Renewable Energy by State (2023)",
    fill = "% Renewable",
    caption = "Data source: U.S. State Renewable Energy Dataset created by Sievers"
  ) +
  theme_minimal()
energy_2023 <- project_long %>%
  filter(year == 2023) %>%
  mutate(
    EVs = as.numeric(e_vs),
    total_renewable_2023 = as.numeric(total_renewable_energy),
    renewable_per_ev = total_renewable_2023 / EVs
  )
summary(energy_2023$renewable_per_ev)
states_sf <- states(cb = TRUE) %>% filter(!STUSPS %in% c("GU","AS","PR","VI","MP"))

states_merged <- states_sf %>%
  left_join(energy_2023, by = c("NAME" = "state"))


## This code uses Leaflet package to create an interactive plot
pal <- colorNumeric("YlGnBu", domain = states_merged$pct_renewable)
leaflet(states_merged) %>%
  addProviderTiles("CartoDB.Positron") %>%
  addPolygons(
    fillColor = ~pal(pct_renewable),
    weight = 1,
    opacity = 1,
    color = "white",
    dashArray = "3",
    fillOpacity = 0.8,
    label = ~paste0(NAME, ": ", round(pct_renewable, 1), "% Renewable"),
    highlight = highlightOptions(
      weight = 2,
      color = "#666",
      fillOpacity = 0.9,
      bringToFront = TRUE
    )
  ) %>%
  addLegend(
    pal = pal,
    values = ~pct_renewable,
    opacity = 0.7,
    title = "Percent Renewable",
    position = "bottomright"
  )
states_sf <- tigris::states(cb = TRUE) %>%
  st_as_sf() %>%
  filter(!STUSPS %in% c("GU", "AS", "PR", "VI", "MP"))
states_merged <- states_sf %>%
  left_join(energy_2023, by = c("NAME" = "state"))
energy_2023 <- project_long %>% 
  filter(year == 2023) %>%
  mutate(
    state_name = str_to_title(state),
    renewable_per_ev = total_renewable_energy / e_vs  
  )

# This code merges with the states Shapefiles
states_merged <- states_sf %>%
  left_join(energy_2023, by = c("NAME" = "state_name"))
# This code creates a color palette
pal_ev <- colorNumeric(
  palette = "YlOrRd",
  domain = states_merged$renewable_per_ev,
  na.color = "grey90"
)
leaflet(states_merged) %>%
  addProviderTiles("CartoDB.Positron") %>%
  addPolygons(
    fillColor = ~pal_ev(renewable_per_ev),
    weight = 1,
    opacity = 1,
    color = "white",
    dashArray = "3",
    fillOpacity = 0.8,
    label = ~paste0(NAME, ": ", round(renewable_per_ev, 1), " units per EV"),
    highlight = highlightOptions(
      weight = 2,
      color = "#666",
      fillOpacity = 0.9,
      bringToFront = TRUE
    )
  ) %>%
  addLegend(
    pal = pal_ev,
    values = ~renewable_per_ev,
    opacity = 0.7,
    title = "Renewable Energy per EV",
    position = "bottomright"
  )
project_long <- Project4 %>%
  pivot_longer(
    cols = c(renewable_use_2021:renewable_use_2023,
             total_renewable_energy_2021:total_renewable_energy_2023,
             av_price_2021:av_price_2023),
    names_to = c(".value", "year"),
    names_pattern = "(.*)_(\\d+)"
  ) %>%
  mutate(
    year = as.numeric(year),
    e_vs = as.numeric(e_vs),  # ensure EVs are numeric
    pct_renewable = (renewable_use / total_renewable_energy) * 100,
    ev_per_total_energy = total_renewable_energy / e_vs
  )
## LOAD SHAPEFILES AND INCLUDE ALL 50 STATES
states_sf <- tigris::states(cb = TRUE) %>%
  st_as_sf() %>%
  filter(STUSPS %in% c(
    "AL","AK","AZ","AR","CA","CO","CT","DE","FL","GA","HI","ID","IL","IN","IA",
    "KS","KY","LA","ME","MD","MA","MI","MN","MS","MO","MT","NE","NV","NH","NJ",
    "NM","NY","NC","ND","OH","OK","OR","PA","RI","SC","SD","TN","TX","UT","VT",
    "VA","WA","WV","WI","WY","DC"
  ))
ggplot(states_merged) +
  geom_sf(aes(fill = total_renewable_energy)) +
  scale_fill_viridis_c(option = "plasma", na.value = "grey90") +
  facet_wrap(~year) +  # one map per year
  labs(
    title = "Total Renewable Energy by State (2021–2023)",
    fill = "Renewable Energy",
    caption = "Data source: Project 4"
  ) +
  theme_minimal()


```
